\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-02-10 paper.tex
Generate the true model for s-curve

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

To investigate the effect of removing long edges on the MSE, we analyzed
the MSE for various benchmark values. Surprisingly, the results
indicated that the MSE remained consistent across different benchmark
values, as shown in Figure~\ref{fig-diagnosticpltScurvelgrm}. It's
important to note that edges are defined only in the 2D model and do not
extend to higher dimensions. Consequently, removing edges does not
affect the model in higher dimensions.

There is no definitive rule for determining the benchmark value to
remove long edges. However, we have a method to find a default value
(see Appendix). Adjusting values around this default can help to find
benchmark value to remove long edges, contributing to the construction
of a smoother 2D representation (see
Figure~\ref{fig-diagnosticpltScurvelgrm}).

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvelgrm-1.pdf}

\caption{\label{fig-diagnosticpltScurvelgrm}Goodness of fit statistics
from UMAP applied to training S-curve dataset with different benchmark
values to remove the long edges. What is the effective benchmark value
to remove the long edges? \(1.347\) is the chosen benchmark value to
remove long edges. The value is selected through a visual inspection of
the triangular mesh is conducted to verify whether it successfully
eliminates long edges.}

@@

The approach involves employing the K-nearest neighbors (KNN) algorithm
to identify the nearest hexagonal bin centroid in the 2D space.
Subsequently, the coordinates of this centroid are assigned as the
low-dimensional predicted values for the test data in 2D space. It is
noteworthy that traditional NLDR methods, such as tSNE, often lack a
direct predict function, making our approach valuable for generating
predicted values in the absence of such functionalities.

@@

Here, \(n\) represents the number of bins, \(p\) represents the
dimensions, \(x_{ij}\) is the actual high-dimensional data, and
\(C_{x_ij}\) is the predicted bin centroid data in high dimensions.

@@

\hypertarget{model-interpretation}{%
\subsection{Model interpretation}\label{model-interpretation}}

The visualizations of NLDR techniques for the S-curve dataset shows
different representations in 2D, as showcased in Figure
Figure~\ref{fig-nldervis}. The question at hand is determining the best
NLDR technique to represent the S-curve data. Examining the 2D layouts
in Figure~\ref{fig-nldervis} suggests that TriMAP or PaCMAP may be best
representations. However, the challenge is to decide the best
representation with considerable evidence.

After considering the models constructed from the algorithm for each
NLDR techniques and visualizing them in high-D space with the original
data, shows that all the NLDR techniques capture the non-linearity of
the S-curve (see videos of the models linked in
\textbf{?@fig-modelScurve}). But thorough examination of each model in
high-D space shows some surprising findings. The model of PHATE is
flattened and lack of capturing the width of S-curve (see videos of the
model with PHATE linked in \textbf{?@fig-modelScurve}). This nature of
PHATE is also evicted by the considerably higher AIC and MSE values for
PHATE as shown in Figure~\ref{fig-diagnosticpltScurve}. Furthermore,
there are two edges that are not connected to any points in the model of
TriMAP (see see videos of the model with TriMAP linked in
\textbf{?@fig-modelScurve}). This model shows how the patterns in 2D
representation of TriMAP capture by the model (see
Figure~\ref{fig-nldervis} (d)). As shown in Figure~\ref{fig-nldervis}
(b), UMAP have four clusters in a curvilinear structure. However, the
model reveals that even though UMAP shows a curvilinear in 2D, UMAP
actually capture the structure of the S-curve without any disjoint in
the model (see see videos of the model with UMAP linked in
\textbf{?@fig-modelScurve}). In the 2D layouts of tSNE and PaCMAP (see
Figure~\ref{fig-nldervis} (a), (e)), there are some sparse areas that
are captured by the models effectively (see see videos of the model with
tSNE and PaCMAP linked in \textbf{?@fig-modelScurve}).

\hypertarget{sec-prediction}{%
\subsection{Prediction}\label{sec-prediction}}

In this context, ``prediction'' denotes the 2D embedding generated for
the NLDR technique. The methodology encompasses identifying the nearest
averaged high-D points in the high-D space for the test data, by
computing high-D Euclidean distances. As the averaged high-D point
corresponds to the lifting of the 2D model, determining its nearest
counterpart allows us to map its hexagonal bin centroid coordinates.
Consequently, these centroid coordinates serve as the assigned
low-dimensional predicted values for the test data in the 2D space.

Some NLDR techniques, such as tSNE, often lack a direct prediction,
making our approach valuable for generating predicted values in the
absence of such functionalities.

@@

\begin{itemize}
\item
  Explain what ``Evaluate Festem on DEG detection'' means, in terms of
  this data
\item
\item
  How did they ``identified 10 clusters''? What were the variables used,
  PCs or full data or 2D representation, spell this out.
\item
  Used top 1000 genes selected by Festem to run PCA, top 15 PCs
  (original paper mentioned)
\item
  The ``10 clusters included immune cells such as naive CD4 T \ldots{}''
  does this mean that every cluster had all of these, or one cluster was
  mostly ``CD4 T cells'', another mostly ``memory CD4 T'', \ldots?
\item
  How is your use of this data different from the original application.
\item
  Original paper used this data set to compare different DEG methods,
  but we need to look how the performance of UMAP with the author's
  selected parameter choice.
\item
  Why 15 PCs? Is that from the original paper? YES
\end{itemize}

@@

@@

@@

@@

@@

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

To investigate the effect of removing long edges on the MSE, we analyzed
the MSE for various benchmark values. Surprisingly, the results
indicated that the MSE remained consistent across different benchmark
values, as shown in \textbf{?@fig-diagnosticpltScurvelgrm}. It's
important to note that edges are defined only in the 2D model and do not
extend to higher dimensions. Consequently, removing edges does not
affect the model in higher dimensions.

There is no definitive rule for determining the benchmark value to
remove long edges. However, we have a method to find a default value
(see Appendix). Adjusting values around this default can help to find
benchmark value to remove long edges, contributing to the construction
of a smoother 2D representation (see
\textbf{?@fig-diagnosticpltScurvelgrm}).

@@

@@

The prediction approach involves employing the K-nearest neighbors (KNN)
algorithm. In this method, which operates as an unsupervised
classification problem, the nearest high-D model point (averaged high-D
point) is identified for the new high-D data point. Once the nearest
high-D model point is identified, its coordinates are used as the
predicted 2D embeddings for the new high-D data point. This step is
particularly valuable due to the limitations of certain NLDR techniques,
like tSNE, which don't provide a straightforward method for predicting
2D embeddings. As a result, our approach presents a solution that can
generate predicted 2D embeddings, irrespective of the NLDR technique
employed, effectively addressing this gap in functionality.

@@

Here, \(n\) represents the number of observations, \(p\) represents the
dimensions, \(x_{ij}\) is the actual high-dimensional data, and
\(C_{x_ij}\) is the predicted bin centroid data in high dimensions.

@@

@@

\begin{itemize}
\item
  Explain what ``Evaluate Festem on DEG detection'' means, in terms of
  this data
\item
  How did they ``identified 10 clusters''? What were the variables used,
  PCs or full data or 2D representation, spell this out.
\item
  Used top 1000 genes selected by Festem to run PCA, top 15 PCs
  (original paper mentioned)
\item
  The ``10 clusters included immune cells such as naive CD4 T \ldots{}''
  does this mean that every cluster had all of these, or one cluster was
  mostly ``CD4 T cells'', another mostly ``memory CD4 T'', \ldots?
\item
  How is your use of this data different from the original application.
\item
  Original paper used this data set to compare different DEG methods,
  but we need to look how the performance of UMAP with the author's
  selected parameter choice.
\item
  Why 15 PCs? Is that from the original paper? YES
\end{itemize}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC-1.pdf}

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC2-1.pdf}

\caption{\label{fig-diagnosticpltPBMC2}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC3-1.pdf}

\caption{\label{fig-diagnosticpltPBMC3}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltHARPS-1.pdf}

\caption{\label{fig-diagnosticpltHARPS}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltHARPS2-1.pdf}

\caption{\label{fig-diagnosticpltHARPS2}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltHARPS3-1.pdf}

\caption{\label{fig-diagnosticpltHARPS3}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}
